The ChEMBL database as Linked Open Data

### 2012-04-15 article.tex
rearrange Results section and edit some parts

Although many parts of the CHEMBL-RDF dataset use community based ontologies, some 
terms from the previous ontology are still used. These terms were created under the 
URI namespace \textit{http://rdf.farmbio.uu.se/chembl/onto/\#}, and are referenced here using the
prefix: \textit{chembl}.

@@

We here present the updated ChEMBL-RDF.

@@

\subsubsection{Added Data}

RSC's ChemSpider~\cite{Pence2010} is a free online database of over 26 million unique
chemical compounds aggregated from over 400 data sources as well as chemical data extracted
from RSC scientific articles and databases. Since its inception efforts have been made to
utilize the platform as both a deposition platform for the community to contribute data as
well as a platform for annotation and curation. Studies have shown that there are data
quality issues in many of the public compound databases~\cite{Williams2011} and ChemSpider has become a
valuable resource for curated data, especially chemical-compound name mappings. ChemSpider
is presently providing the chemical services underpinning the Open PHACTS semantic web
project~\cite{Williams2012} providing access to structure, substructure and similarity searching services
to the core architecture of the project. Specific chemical data sources containing data
mappings between ChemSpider identifiers (CSIDs) and the original data source identifiers
have been provided to the triple store, together with chemical identifiers including
validated chemical names (systematic, generic and trivial), SMILES () and InChIs (). 
The data mappings between the ChemSpider IDs and the data source IDs are released to
the community under Creative Commons licenses (CC-BY-SA 3.0; attribution should be
made to the original ChEMBL database, Open PHACTS, and ChemSpider).

@@

The core concept in the ChEMBL database is that of the biological activity. This
is the type of information that the ChEMBL database extracts from literature.
Activities are at this moment still using an informal, custom vocabulary: the
type, fields and links to other resources are using the \textit{chembl}
namespace. While ChEMBL provides both the original activities as found in literature
and standardized values allowing comparison between studies, the triples only
make the latter available. The CiTO ontology is used to point to the paper from
which the data was extracted.

@@

@@

Besides the PubMed identifiers used to link from literature references to Bio2RDF (see earlier),
ChEMBL uses DOIs alternatively, which we use to link out to the RDF provided by CrossRef~\citep{Bilder2011}:

@@

The ChEMBL-RDF is increasingly used, and we here present a few applications. The first application
describes how it uses the SPARQL end point can be used to make the data available as linked data
via the Bio2RDF platform. The second application takes advantage of the bibliographic information
exposed as machine readable data, in calculating citation statistics. The third application shows
an integration of ChEMBL-RDF with ChemSpider to provide an extension for the decision support
platform in Bioclipse. The last application shows how Chem2Bio2RDF can combine the ChEMBL data
with other life sciences databases, showing the power of the Linked Data approach.

@@

The Linked Data server used by Bio2RDF has been reconfigured for ChEMBL to provide URL
based services for standard URI resolution, along with text and link searches
\cite{WebAppGitHub}. It proxies the standard item
URIs by translating URLs between those requested by users and the URIs that are available
in the databases. For example, if the ChEMBL web application is running on the users
local machine, e.g. \url{http://localhost:8080/chembl/}, then a request for the
article with identifier ``a31863'', \url{http://localhost:8080/-chembl/article/a31863},
will be resolved from the database using the full original URI,
\url{http://data.kasabi.com/-dataset/chembl-rdf/13/activity/a31863}. If the
user requested an RDF document, using content negotiation, the original URIs will be unchanged,
however, if the user requested an HTML document, the results will contain both the
original RDF triples, represented using RDFa, with links that resolve using the users local machine.

@@

In this study we extended Bioclipse DS with remote access to ChEMBL-RDF and Chemspider, which enables users to for chemical structures in Bioclipse look up near neighbors on Chemspider via the Chemspider Web API (SOAP) and query ChEMBL-RDF for properties regarding any measured interactions. The results are presented alongside predictive models in Bioclipse, and can be used as decision support when evaluating chemical structures and consider strategies for e.g. optimization.

>>>>

Although many parts of the CHEMBL-RDF 13 dataset now use community based ontologies, some 
terms from the previous ad-hoc ontology are still used. These terms were created under the 
URI namespace \textit{http://rdf.farmbio.uu.se/chembl/onto/\#}, and are referenced here using the
prefix: \textit{chembl}.

@@

We here present the updated ChEMBL-RDF, including a description of the mapping process 
and the identification of links to other datasets.

@@

@@

The core concept in the ChEMBL database is that of the biological activity. This
is the type of information that the ChEMBL database extracts from literature.
In CHEMBL-RDF activities are mostly represented using the custom CHEMBL-RDF vocabulary, where the
type, fields and links to other resources are using the \textit{chembl}
namespace. While ChEMBL provides both the original activities as found in literature
and standardized values allowing comparison between studies, the triples only
make the latter available. The CiTO ontology is used to point to the paper from
which the data was extracted.

@@

ChemSpider~\cite{Pence2010} is a free online database provided by RSC. It contains over 26 million unique
chemical compounds aggregated from over 400 data sources as well as chemical data extracted
from RSC scientific articles and databases. Since its inception, efforts have been made to
utilize it as both a deposition platform for the community to contribute novel data, as
well as a platform for annotation and curation for existing data. Studies have shown that there are data
quality issues in many of the public compound databases~\cite{Williams2011}. ChemSpider has become a
valuable resource for curated data, especially chemical-compound name mappings. ChemSpider
is presently providing the chemical structure, substructure and similarity searching services 
underpinning the Open PHACTS semantic web project~\cite{Williams2012}. Specific chemical data 
sources containing data mappings between ChemSpider identifiers (CSIDs) and the original data 
source identifiers have been provided to the triple store, together with chemical identifiers 
including validated chemical names (systematic, generic and trivial), SMILES () and InChIs (). 
The data mappings between the CSIDs and ChEMBL IDs are released to the 
community under the Creative Common Attribution-Share Alike license (CC-BY-SA 3.0). Attribution 
should be made to the original ChEMBL database, Open PHACTS, and ChemSpider.

@@

In addition to the PubMed identifiers used to link from literature references to Bio2RDF,
ChEMBL provides DOIs, which we use to link out to the RDF provided by CrossRef~\citep{Bilder2011}:

@@

ChEMBL-RDF can be used to explore interrelated scientific datasets, and we here present a few applications. 
The first application describes how it uses the SPARQL end point can be used to make the data available as linked data
via the Bio2RDF platform. The second application takes advantage of the bibliographic information
exposed as machine readable data, in calculating citation statistics. The third application shows
an integration of ChEMBL-RDF with ChemSpider to provide an extension for the decision support
platform in Bioclipse. The last application shows how Chem2Bio2RDF can combine the ChEMBL data
with other life sciences databases, showing the power of the Linked Data approach.

@@

The Linked Data server used by Bio2RDF has been reconfigured for ChEMBL to provide URL
based services for standard URI resolution, along with text and link searches~\cite{Ansell2011}. 
The configuration files along with build scripts for the ChEMBL-RDF Linked Data server are available on GitHub~\cite{WebAppGitHub}. 
It proxies the standard item URIs by translating URLs between those requested by users and the URIs that are available
in SPARQL endpoints. For example, if the ChEMBL web application is running on the users
local machine, e.g. \url{http://localhost:8080/chembl/}, then a request for the
article with identifier ``a31863'', \url{http://localhost:8080/-chembl/article/a31863},
will be resolved from the database using the full original URI,
\url{http://data.kasabi.com/-dataset/chembl-rdf/13/activity/a31863}. If the
user requested an RDF document using content negotiation, the original URIs will be unchanged,
however, if the user requested an HTML document, the results will contain both the
original RDF triples, represented using RDFa, with links that resolve using the users local machine.

@@

In this study we extended Bioclipse DS with remote access to ChEMBL-RDF and Chemspider. 
This enables users browsing chemical structures in Bioclipse to look up near neighbors on Chemspider 
via the Chemspider Web API (SOAP) and query ChEMBL-RDF for properties regarding any measured interactions. 
The results are presented alongside predictive models in Bioclipse, and can be used as 
decision support when evaluating chemical structures and consider strategies for optimization.
